---
title: "Genomic regions from Ensembl and UCSC"
author: '[Benjamin Leblanc](https://github.com/benja0x40)'
date: "`r format(Sys.time(), '%d.%m.%Y')`"
vignette: >
  %\VignetteIndexEntry{Genomic regions from Ensembl and UCSC}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
fontsize: 11pt
output:
  html_document:
    number_sections: true
    toc: true
    toc_depth: 2
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r hidden_dependencies, include=FALSE}
library(knitr)
suppressPackageStartupMessages(library(Tightrope))
```

___

The purpose of this documentation is to provide a practical guide to the use of
`R` functions available in `Tightrope`. A basic knowledge of the `R` syntax
and data structures, and of [Bioconductor](http://www.bioconductor.org/)
objects used to represent and manipulate mapped reads and genomic intervals
will help to understand and adapt the included source code examples to other
ChIP-seq datasets.

___

\newpage

# Summary

# Gene features from Ensembl

## Provided regions

| Organism     | Genome build  | Description                        | R data object |
|--------------|---------------|------------------------------------|---------------|
| Homo sapiens | GRCh38 (hg38) | Ensembl genes (rel. 91, Dec. 2017) | EGA91_human   |
| Homo sapiens | GRCh37 (hg19) | Ensembl genes (rel. 75, Feb. 2014) | EGA75_human   |
| Mus musculus | GRCm38 (mm10) | Ensembl genes (rel. 91, Dec. 2017) | EGA91_mouse   |
| Mus musculus | NCBIM37 (mm9) | Ensembl genes (rel. 67, May 2012)  | EGA67_mouse   |

| Organism     | Genome build  | Source                     | R data object |
|--------------|---------------|----------------------------|---------------|
| Homo sapiens | GRCh38 (hg38) | UCSC / cpgIslandExt.txt.gz | CGI_hg38      |
| Homo sapiens | GRCh37 (hg19) | UCSC / cpgIslandExt.txt.gz | CGI_hg19      |
| Mus musculus | GRCm38 (mm10) | UCSC / cpgIslandExt.txt.gz | CGI_mm10      |
| Mus musculus | NCBIM37 (mm9) | UCSC / cpgIslandExt.txt.gz | CGI_mm9       |

```{r}
# Mouse genome annotations (Ensembl GRCm38.p5 release 91)
data("EGA91_mouse")    # Gene features from Ensembl
data("CGI_mm10")       # CpG Islands from UCSC
data("mm10_blacklist") # Blacklisted regions from ENCODE
```

## Making regions

```{r}

```

# Importing regions





