---
title: "BRD normalization"
author: "Benjamin Leblanc"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Analysis template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Helper functions

```{r, include=FALSE}
# =============================================================================.
#
# -----------------------------------------------------------------------------.
ExperimentalAnnotations <- function(wks_name, xps = T) {
  tbl <- get(wks_name)$ann[xps, ]
  if(any(grepl("^geo\\.", colnames(tbl)))) {
    skip <- c(
      "experiment", "submission", "study", "sample", "run", "molecule"
    )
  } else {
    skip <- c(
      "project", "SampleNumber", "IsPairedEnd", "Read1", "Read2", "NumReadsPF"
    )
  }
  skip <- c(
    skip, "mapping_command", "basemount.host", "mapping_options", "mapping_index"
  )
  rex <- "^(geo|library_|platform_|instrument_)|(file|path)"
  skip <- colnames(tbl) %in% skip | grepl(rex, colnames(tbl))
  tbl <- tbl[, ! skip]
  tbl
}
# =============================================================================.
#
# -----------------------------------------------------------------------------.
InsertAntibody <- function(viz, abd) {
  for(i in 1:length(viz)) {
    viz[[i]] <- gsub("X---X", abd, viz[[i]])
  }
  viz
}
```

### Initializations

```{r, include=FALSE}
# #############################################################################.
suppressPackageStartupMessages(library(Tightrope))
# =============================================================================.
#
# -----------------------------------------------------------------------------.
CFG_DTS <- list()
# -----------------------------------------------------------------------------.
# ESC_BRD, NSC_K27M, E14_K27M, Lu_et_al
load_dataset   <- "ESC_BRD"
target_dataset <- "ESC_BRD"
antibody <- "Suz12"
```

### Collection of spike-in datasets

```{r, include=FALSE}
# =============================================================================.
if(any(c("ESC_BRD", "ALL") %in% load_dataset)) {
  ESC_BRD <- list()
  
  data_path <- "/Volumes/USB16GB/LT_WORKS/NextSeq_ESC_BRD_R1_TEST/"
  ESC_BRD$R1 <- list(
    ann = read.delim(
      paste0(data_path, "_LittleThumb_/datasets/bowtie2_mm10.txt"),
      comment.char = "#", stringsAsFactors = F
    ),
    CNT = readRDS(paste0(data_path, "_RDATA_/ROI.CNT.rdata"))
  )
  
  data_path <- "/Volumes/USB16GB/LT_WORKS/NextSeq_ESC_BRD_R2_TEST/"
  ESC_BRD$R2 <- list(
    ann = read.delim(
      paste0(data_path, "_LittleThumb_/datasets/bowtie2_mm10.txt"),
      comment.char = "#", stringsAsFactors = F
    ),
    CNT = readRDS(paste0(data_path, "_RDATA_/ROI.CNT.rdata"))
  )
  rm(data_path)
  
  # Reshape to join replicates
  ESC_BRD <- list(
    name = "ESC_BRD",
    antibody = antibody,
    data_path = "/Volumes/USB16GB/LT_WORKS/NextSeq_ESC_BRD_RDATA/",
    ann = rbind(
      ESC_BRD$R1$ann,
      ESC_BRD$R2$ann
    ),
    CNT = with(ESC_BRD, JoinColumns(R1$CNT, R2$CNT))
  )
  
  if(! with(ESC_BRD, all(ann$lt_id == colnames(CNT$GN)))) {
    stop("Inconsistent annotation of samples in ESC_BRD")
  }
  # ---------------------------------------------------------------------------.
  ESC_BRD <- list2env(ESC_BRD)
}

# =============================================================================.
if(any(c("NSC_K27M", "ALL") %in% load_dataset)) {
  data_path <- "/Volumes/USB16GB/LT_WORKS/NextSeq_K27M_spike_in/"
  NSC_K27M <- list(
    name = "NSC_K27M",
    antibody = antibody,
    data_path = paste0(data_path, "_RDATA_/"),
    ann = read.delim(
      paste0(data_path, "_LittleThumb_/datasets/bowtie2_mm10.txt"),
      comment.char = "#", stringsAsFactors = F
    ),
    CNT = readRDS(paste0(data_path, "_RDATA_/ROI.CNT.rdata"))
  )
  rm(data_path)
  
  if(! with(NSC_K27M, all(ann$lt_id == colnames(CNT$GN)))) {
    stop("Inconsistent annotation of samples in NSC_K27M")
  }
  # ---------------------------------------------------------------------------.
  NSC_K27M <- list2env(NSC_K27M)
}

# =============================================================================.
if(any(c("E14_EPZ", "ALL") %in% load_dataset)) {
  data_path <- "/Volumes/USB16GB/LT_WORKS/NextSeq_E14EPZ/"
  E14_EPZ <- list(
    name = "E14_EPZ",
    antibody = antibody,
    data_path = paste0(data_path, "_RDATA_/"),
    ann = read.delim(
      paste0(data_path, "_LittleThumb_/datasets/bowtie2_mm10.txt"),
      comment.char = "#", stringsAsFactors = F
    ),
    CNT = readRDS(paste0(data_path, "_RDATA_/ROI.CNT.rdata"))
  )
  rm(data_path)
  
  if(! with(E14_EPZ, all(ann$lt_id == colnames(CNT$GN)))) {
    stop("Inconsistent annotation of samples in E14_EPZ")
  }
  # ---------------------------------------------------------------------------.
  E14_EPZ <- list2env(E14_EPZ)
}

# =============================================================================.
if(any(c("Lu_et_al", "ALL") %in% load_dataset)) {
  data_path <- "/Volumes/USB16GB/LT_WORKS/GSE63195_Lu_et_al_2016/"
  Lu_et_al <- list(
    name = "Lu_et_al",
    antibody = antibody,
    data_path = paste0(data_path, "_RDATA_/"),
    ann = read.delim(
      paste0(data_path, "_LittleThumb_/datasets/bowtie2_mm10.txt"),
      comment.char = "#", stringsAsFactors = F
    ),
    CNT = readRDS(paste0(data_path, "_RDATA_/ROI.CNT.rdata"))
  )
  rm(data_path)
  
  if(! with(Lu_et_al, all(ann$lt_id == colnames(CNT$GN)))) {
    stop("Inconsistent annotation of samples in Lu_et_al")
  }
  # ---------------------------------------------------------------------------.
  Lu_et_al <- list2env(Lu_et_al)
}
```

#### 1. ESC_BRD

SUZ12 WT + KO cell mix experiments in mouse ESCs.

```{r, include=FALSE}
wks_name <- "ESC_BRD"
# =============================================================================.
if(exists(wks_name)) {
  wks <- get(wks_name)
  CFG_DTS[[wks_name]] <- list(
    xps = wks$ann$antibody %in% c(antibody, "Input"), # & replicate %in% c("R1", "R2")
    viz = list(
      c("ESC_Input_0_R1", "ESC_X---X_100_R1", "ESC_X---X_10_R1", "ESC_X---X_0_R1"),
      c("ESC_Input_0_R2", "ESC_X---X_100_R2", "ESC_X---X_10_R2", "ESC_X---X_0_R2"),
      
      c("ESC_X---X_100_R1", "ESC_X---X_10_R1"),
      c("ESC_X---X_100_R2", "ESC_X---X_10_R2"),
      c("ESC_Input_0_R1",      "ESC_Input_0_R2")
    ),
    layout = matrix(1:9, 3, 3, byrow = T),
    nxp  = 0,
    rex  = "Input",
    cvt  = 0.5,
    npc  = NA,
    knn  = 200,
    cst  = 0.5,
    ncl  = 2
  )
  CFG_DTS[[wks_name]]$viz <- InsertAntibody(CFG_DTS[[wks_name]]$viz, antibody)
  CFG_DTS[[wks_name]]$nxp <- sum(CFG_DTS[[wks_name]]$xps)
  # ---------------------------------------------------------------------------.
  rm(wks)
}
```

```{r, echo=FALSE, results='asis'}
if(exists(wks_name)) {
  knitr::kable(ExperimentalAnnotations(wks_name, xps = CFG_DTS[[wks_name]]$xps))
  rm(wks_name)
}
```

#### 2. NSC_K27M

H3WT versus K27M experiments in mouse NSCs.

```{r, include=FALSE}
wks_name <- "NSC_K27M"
# =============================================================================.
if(exists(wks_name)) {
  wks <- get(wks_name)
  CFG_DTS[[wks_name]] <- list(
    xps = with(
      wks$ann, antibody %in% c("H3K27me3", "Input") & replicate %in% c("R1", "R2")
    ),
    viz = list(
      c("H3WT_DMSO_H3K27me3_R1", "K27M_DMSO_H3K27me3_R1"),
      c("H3WT_DMSO_Input_R1",    "H3WT_DMSO_H3K27me3_R1", "K27M_DMSO_H3K27me3_R1"),
      
      c("H3WT_DMSO_H3K27me3_R2", "K27M_DMSO_H3K27me3_R2"),
      c("H3WT_DMSO_Input_R2",    "H3WT_DMSO_H3K27me3_R2", "K27M_DMSO_H3K27me3_R2"),
      
      c("H3WT_DMSO_H3K27me3_R1", "H3WT_DMSO_H3K27me3_R2"),
      c("K27M_DMSO_H3K27me3_R1", "K27M_DMSO_H3K27me3_R2"),
      c("H3WT_DMSO_Input_R1",    "H3WT_DMSO_Input_R2")
    ),
    layout = matrix(1:9, 3, 3, byrow = T),
    nxp  = 0,
    rex  = "Input",
    cvt  = 0.5,
    npc  = NA,
    knn  = 300,
    cst  = 0.9,
    ncl  = 1
  )
  CFG_DTS[[wks_name]]$nxp <- sum(CFG_DTS[[wks_name]]$xps)
  # ---------------------------------------------------------------------------.
  rm(wks)
}
```

```{r, echo=FALSE, results='asis'}
if(exists(wks_name)) {
  knitr::kable(ExperimentalAnnotations(wks_name, xps = CFG_DTS[[wks_name]]$xps))
}
rm(wks_name)
```

#### 3. E14_EPZ

EZH2 inhibitor washout experiments in mouse ESCs.

```{r, include=FALSE}
wks_name <- "E14_EPZ"
# =============================================================================.
if(exists(wks_name)) {
  wks <- get(wks_name)
  CFG_DTS[[wks_name]] <- list(
    xps = wks$ann$antibody %in% antibody,
    viz = list(
      c("X---X_E14", "X---X_7d", "X---X_8h", "X---X_24h")
    ),
    layout = matrix(1:9, 3, 3, byrow = T),
    nxp  = 0,
    rex  = c(H3K27me3 = "_7d$", H3K27me2 = "_7d$", H3K27me1 = "_7d$"),
    cvt  = 0.5,
    npc  = NA,
    knn  = 300,
    cst  = c(H3K27me3 = .95, H3K27me2 = 0.5, H3K27me1 = 0.5),
    ncl  = c(H3K27me3 =   1, H3K27me2 =   2, H3K27me1 =   3)
  )
  CFG_DTS[[wks_name]]$rex  <- CFG_DTS[[wks_name]]$rex[antibody]
  CFG_DTS[[wks_name]]$cst  <- CFG_DTS[[wks_name]]$cst[antibody]
  CFG_DTS[[wks_name]]$ncl  <- CFG_DTS[[wks_name]]$ncl[antibody]
  
  CFG_DTS[[wks_name]]$viz <- InsertAntibody(CFG_DTS[[wks_name]]$viz, antibody)
  CFG_DTS[[wks_name]]$nxp <- sum(CFG_DTS[[wks_name]]$xps)
  # ---------------------------------------------------------------------------.
  rm(wks)
}
```

```{r, echo=FALSE, results='asis'}
if(exists(wks_name)) {
  knitr::kable(ExperimentalAnnotations(wks_name, xps = CFG_DTS[[wks_name]]$xps))
}
rm(wks_name)
```

#### 4. Lu_et_al

H3WT versus K36M experiments from Lu et al., 2016.

```{r, include=FALSE}
wks_name <- "Lu_et_al"
# =============================================================================.
if(exists(wks_name)) {
  wks <- get(wks_name)
  CFG_DTS[[wks_name]] <- list(
    xps = wks$ann$antibody %in% c(antibody, "Input"),
    viz = list(
      c("H3WT_X---X", "K36M_X---X"),
      c("H3WT_Input", "H3WT_X---X", "K36M_X---X")
    ),
    layout = matrix(1:9, 3, 3, byrow = T),
    nxp  = 0,
    rex  = "Input",
    cvt  = 0,
    npc  = 3, # Force to conserve 3 dimensions of variation: H3WT, K36M, Input
    knn  = 300,
    cst  = 0.7,
    ncl  = 2
  )
  CFG_DTS[[wks_name]]$viz <- InsertAntibody(CFG_DTS[[wks_name]]$viz, antibody)
  CFG_DTS[[wks_name]]$nxp <- sum(CFG_DTS[[wks_name]]$xps)
  # ---------------------------------------------------------------------------.
  rm(wks)
}
```

```{r, echo=FALSE, results='asis'}
if(exists(wks_name)) {
  knitr::kable(ExperimentalAnnotations(wks_name, xps = CFG_DTS[[wks_name]]$xps))
}
rm(wks_name)
```

### BRD procedure

#### Glopal settings

```{r}
ref_roi <- "GN"
dither  <- 3
smobs   <- T
zscore  <- T
mincs   <- 300
```

#### Dataset dependent settings

```{r, echo=FALSE, results='asis'}
tbl <- t(sapply(CFG_DTS, "[", -(1:3)))
knitr::kable(tbl)
rm(tbl)
```

#### Extraction of count data

```{r, include=FALSE}
# Select dataset to be analysed
cfg  <- CFG_DTS[[target_dataset]]

viz  <- cfg$viz
lyt  <- cfg$layout

# Raw counts
wks <- get(target_dataset)
cnt <- with(wks, CNT[[ref_roi]][, cfg$xps])

# Experiment and controls
xps <- colnames(cnt)
inp <- which(grepl(cfg$rex, xps, perl = T))

# Dithered and log2 transformed counts
ldc <- log2(DitherCounts(cnt))         

# Quick and dirty
cleanup <- FiniteValues(log(cnt))
cnt <- cnt[cleanup, ]
ldc <- ldc[cleanup, ]

rm(cfg, wks, chk)
```

#### BRD estimation

```{r, include=FALSE}

brd <- with(
  cfg, BRD(
    cnt, controls = , dither = dither,
    smobs = smobs, cvt  = cvt , npc = npc, zscore = zscore,
    knn = knn, cst = cst, ncl = ncl, mincs = mincs
  ) 
)

if(F) {
  # Count density after dithering and dimensionality reduction
  dr <- with(
    cfg, cdadadr(
      cnt, knn = knn, smobs = smobs, cvt  = cvt , npc = npc, zscore = zscore,
      dither = dither
    ) 
  )

  xr <- list(x = dr$projection$x, d = dr$density)
  
  # Moderate density threshold to increase clustering robustness
  h <- which(rankstat(xr$d) > cfg$cst )
  h <- list(i = h, x = xr$x[h, ], d = xr$d[h])
  
  # Clustering by density gradient ascent
  qs <- QuickShiftClustering(h$x, h$d, n = cfg$ncl)
  
  # Find background cluster
  if(ncl > 1) {
    grp <- bg_grp(ldc[h$i, ], grp = qs$membership, bg_cols = inp)
  } else {
    grp <- 1
  }
  bg <- h$i[qs$membership == grp[1]]
  
  # Fit multivariate gaussian model to the background cluster
  mm <- mv_gmm(ldc[bg, ], ns = 1)
  bg_theta <- mm$theta[[1]]
}
```

#### Save scaling factors

```{r}
wks <- get(target_dataset)
wks$experiments <- xps
wks$CNT <- ExtractColumns(wks$CNT, xps)
wks$BRD <- list(
  ref_roi  = ref_roi,           # regions used for BRD
  control  = xps[inp],          # samples used as Input/IgG
  cleanup  = cleanup,           # observation with count > 0 in ref_roi data
  drstatus = dr$projection$msg, # status of dimensionality reduction with cdadadr
  reduced  = xr,                # dimensionality reduced count data 
  coreset  = h,                 # observations with density above threshold
  clusters = qs,                # clustering
  ctrlgrps = grp,               # cluster ids ordered by decreasing control / ChIP averages
  bg_theta = bg_theta,          # mv parameters of clusters (gmm fit using EM)
  scaling  = mean(bg_theta$mu) - bg_theta$mu,
  bg_index = bg
)
wks$CFG <- CFG_DTS[[target_dataset]]

saveRDS(wks, paste0(wks$data_path, target_dataset, "_", antibody, ".RData"))
```

#### Results visualization

##### Count density after dithering and dimensionality reduction

```{r, echo=FALSE, results='asis'}
tbl <- t(sapply(dr$projection$msg, rbind))
knitr::kable(data.frame(tbl))
```

```{r, echo=FALSE, results='asis'}
tbl <- data.frame(
  cluster_size = qs$csize,
  background   = 1:ncl == grp[1]
)
knitr::kable(tbl)
```


```{r, echo=FALSE, fig.width=6, fig.height=3.5}
# layout 1 x 2 => 5.28 x 2.97
# layout 2 x 2 => 5.28 x 5.94
# layout 6 x 2 => 5.28 x 8.91

grp.clr <- rgb(1, 1, 0)
if(ncl > 1) {
  grp.clr <- c(
    grp.clr, transformColors(SuperRainbow(3), delta.H = 30)[2:ncl]
  )
}
grp.prm <- defineGroups(ids = grp, colors = grp.clr)

clr.prm <- AutoColorParameters("ry")
clr.prm$thresholds <- round(clr.prm$thresholds, 2)

layout(matrix(1:2, 1, 2, byrow = T))
par(mar = c(5, 4, 3, 1), pch = 20, cex.lab = 0.9, cex.axis = 0.9)

xlim <- range(xr$x[, 1]) * 1.2
ylim <- range(xr$x[, 1:2])
ylim <- ylim * 1.2 - diff(ylim) * 0.1 
plot_samples(xr$x, xlim = xlim, ylim = ylim, col = colorize(xr$d, mode = "rank", colors = "ry"), main = "density")
colorLegend("b", clr.prm, ticks = 0:4/4, horiz = T, size = c(70, 3), tick.pos = -1, cex = 0.75)

plot_samples(xr$x, xlim = xlim, ylim = ylim, col = colorize(xr$d, mode = "rank"), main = "clustering")
plot_samples(xr$x[h$i, ], grp = qs$membership, grp.prm = grp.prm, alpha = 0.1, add = T)
```

##### Selected background population and scaling factors

```{r, echo=FALSE, fig.width=3, fig.height=3.5}
# layout 1 x 1 => 2.64 x 2.97
# layout 1 x 2 => 5.28 x 2.97
# layout 2 x 2 => 5.28 x 5.94

# layout(lyt)
par(mar = c(5, 4, 3, 1), pch = 20, cex.lab = 0.9, cex.axis = 0.9)
for(cmp in viz) {
  for(lbl in cmp[-1]) {
  idx <- match(c(cmp[1], lbl), xps)
  d <- knn_density(ldc[, idx], k = knn)
  plot_samples(log2(cnt), idx, col = colorize(d, mode = "rank"))
  plot_samples(log2(cnt)[bg, ], idx, col = rgb(1, 1, 0), alpha = 0.05, add = T)
  plot_sources_2D(theta = list(bg_theta), components = idx, clr = rgb(1, 0, 0), lwd = 1)
  abline(a = diff(bg_theta$mu[idx]), b = 1, col = rgb(1, 0.5, 0), lwd = 1.5)
  }
}
```

```{r, include=FALSE}
rm(cmp, lbl, idx, d)
```

### END

```{r, eval=FALSE, include=FALSE}

# cgi <- E14_EPZ$CNT$CGI_UCSC
# cgi <- cgi[FiniteValues(log2(cgi)), colnames(nrm)]
# cgi <- log2(DitherCounts(cgi))
# cgi <- t(t(cgi) + scaling)
# 
# layout(matrix(1:2, 1, 2))
# par(mar = c(10, 4, 3, 1), pch = 20, cex.lab = 0.9, cex.axis = 0.9)
# boxplot(nrm, las = 2)
# abline(h = median(nrm[, 1]))
# boxplot(cgi, las = 2)
# abline(h = median(cgi[, 1]))
# 
# o <- order(cgi[, 1], decreasing = T)
# m <- apply(cgi[o, ], 2, runmed, k = 255)
# n <- ncol(m) - 2
# clr <- c("black", rgb(0:n/n, 0.5*n:0/n, n:0/n))
# n <- ncol(m)
# 
# layout(matrix(1:4, 2, 2, byrow = T))
# par(mar = c(5, 4, 3, 1), pch = 20, cex.lab = 0.9, cex.axis = 0.9)
# empty.plot(xlim = c(1, nrow(m)), ylim = range(m))
# for(i in 1:n) points(m[, i], type = "l", col = clr[i])



# me3 <- ldc[, grepl("H3K27me3", colnames(ldc))]
# me2 <- ldc[, grepl("H3K27me2", colnames(ldc))]
# me1 <- ldc[, grepl("H3K27me1", colnames(ldc))]
# 
# x <- me1
# x <- x - rowMeans(x)
# 
# layout(matrix(1:4, 2, 2, byrow = T))
# par(mar = c(5, 4, 3, 1), pch = 20, cex.lab = 0.9, cex.axis = 0.9)
# 
# pcp <- prcomp(x, retx = T, center = T, scale. = T)
# pcp <- pcp$x[, 1:2]
# src <- icafast(x, nc = 2, alg = "def")
# icp <- src$Y[, 1:2]
# src <- src$S[, 1:2]
# 
# pcd <- knn_density(pcp, k = 200)
# icd <- knn_density(icp, k = 200)
# scd <- knn_density(src, k = 200)
# 
# plot_samples(pcp, col = colorize(pcd, mode = "rank", colors = "ry"))
# plot_samples(icp, col = colorize(icd, mode = "rank", colors = "ry"))
# plot_samples(src, col = colorize(scd, mode = "rank", colors = "ry"))


```

